Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-09-28 pnas/Towards.tex
remove highlighted comments and very small modification in the appendix

Figure~\ref{fig:Einst_example} shows several estimated images for a growing number of micrographs, and a movie is available as supplementary material. Figure~\ref{fig:error_per_micro} (see appendix) presents the normalized recovery error as a function of the amount of data available.  This is computed after fixing the reflection symmetries (see Section~\ref{sec:AC_analysis}). As evidenced by these figures, the ground truth image can be estimated increasingly well from increasingly many micrographs, without particle picking.

@@

\subsection{Cryo-EM} The application of autocorrelation analysis to cryo-EM follows the same mathematical principles.
The derivation of the first three autocorrelations of the micrographs and their relations to the volume itself are provided in Appendix~\ref{sec:ac_cryo}.
In particular, numerical evidence suggests that the third-order autocorrelation uniquely determines the 3-D volume. 
Figure~\ref{fig:cryo_recon} shows recovery of the 3-D volume from the clean autocorrelations\revised{ and from noisy micrographs.
Unfortunately, the mapping between the autocorrelations and volume seems to be ill-conditioned, preventing high-resolution recovery from noisy data.}
The details of the reconstruction algorithm are given in Appendix~\ref{sec:numeric_details}.
In the next section, we outline how we suggest to overcome the ill-conditioning in future work.

While we cannot provide a \revised{high-resolution} 3-D reconstruction from noisy data with the current algorithm, our method can be easily applied to the problem of deciding whether a micrograph contains projections or merely pure noise---a problem considered in classical works in statistics~\cite{donoho2004higher} %and debates in the cryo-EM community~\cite{henderson2013avoiding}.
and cryo-EM~\cite{henderson2013avoiding}.
This task can be performed by considering solely the recovered $\gamma$ (the fraction of pixels occupied by projections in the micrograph), estimated as part of the recovery algorithm.
Figure~\ref{fig:cryo_detection} (see appendix) presents excerpts  of two noisy micrographs, only one of which contains projections.
In the presence of projections, the estimated $\gamma$ was $0.12$, corresponding to approximately $6784$ projections. 
On the other hand, the estimated $\gamma$ drops to $10^{-5}$ for the pure noise micrograph, corresponding to less than one projection.

\begin{figure}[h]
	\centering
	\begin{subfigure}[h]{0.45\textwidth}
		\centering
		\includegraphics[angle=270, width=.7\textwidth]{TRPV1_rec_incl_noisy.eps}
		\caption{TPRV1 with cutoff $L_{\text{max}}=5$}\label{fig:TRPV1_recon}
	\end{subfigure}
	\hfill
	\begin{subfigure}[h]{0.45\textwidth}
		\centering
		\includegraphics[angle=270, width=.7\textwidth]{BPTI_maxL2_L20.eps}
		\caption{BPTI with cutoff $L_{\text{max}}=2$}\label{fig:BPTI_recon}
	\end{subfigure} 
	\caption{\label{fig:cryo_recon} Reconstructions from the first, second and third order autocorrelations.
		The ground truth volumes were expanded according to~\eqref{eq:volume_expansion_app} with cutoff $L_{\text{max}}$.
		\revised{The original molecules are shown in purple and the (smoothed) ground truths in blue  to illustrate the smoothing effect of our downsampling and truncation of the spherical harmonics expansion.
		The reconstructions  from clean autocorrelations are shown in yellow and recoveries from 
		autocorrelations estimated from noisy data in gray. For the noisy experiments, we used 300 micrographs with SNRs of 1/16 for TRPV1 and 1/64 for BPTI. We present excerpts from noisy micrographs side-by-side with the corresponding clean ones in Figure~\ref{fig:cryo_noisy_micros_for_rec} (see appendix)}.}
		 \vspace{-20pt} 	
\end{figure}
% -------------- Vertical subfigures: -------------------------------

@@

\begin{figure*}[t!]
	\centering
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{TRPV1_clean_micro.png}
		\caption{}
	\end{subfigure} \hfill %\hspace{2pt}
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{TRPV1_noisy_micro_SNR_16.png}
		\caption{}
	\end{subfigure} \hfill %\hspace{2pt}
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{clean_micro_BPTI.png}
		\caption{}
	\end{subfigure} \hfill
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{noisy_micro_BPTI_SNR_64.png}
		\caption{}
	\end{subfigure}
	\caption{\label{fig:cryo_noisy_micros_for_rec} \revised{Excerpts from the noisy micrographs used for the reconstructions in Figure~\ref{fig:cryo_recon} and the corresponding clean excerpts. (a) Excerpt of the clean TRPV1 micrograph; (b) Excerpt of the noisy TRPV1 micrograph with SNR $= 1/16$; (c) Excerpt from the clean BPTI micrograph; (d) Excerpt from the noisy BPTI micrograph with SNR $= 1/64$.}} 
\end{figure*}

\begin{figure*}[t!]
	\centering
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\includegraphics[scale=0.35]{TRPV1_trunc_FSC.eps}
		\caption{}
	\end{subfigure} \hfill %\hspace{2pt}
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\includegraphics[scale=0.35]{BPTI_FSC.eps}
		\caption{}
	\end{subfigure} 
	\caption{\label{fig:cryo_FSC} \revised{Fourier Shell Correlations (FSCs) for the noisy reconstructions presented in Figure~\ref{fig:cryo_recon} using the $0.5$ criterion, compared to the (smoothed) ground truth. (a) FSC for the TRPV1 reconstruction, giving resolution of $24 \AA$; (b) FSC for the BPTI reconstruction, giving resolution of $13 \AA$.}} 
\end{figure*}

@@

For the \revised{clean} experiments in Figure~\ref{fig:cryo_recon}, we computed the analytical first three autocorrelations of the volume as explained in Appendix~\ref{sec:ac_cryo}.
To estimate the coefficients of the volume itself, we solve the LS problem
\begin{equation} \label{eq:LS_cryo}
\min_{\hat{x},\hat{\gamma}} w_1 \vert \mathfrak{a}_x^1 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^1 \vert^2 + w_2\| \mathfrak{a}_x^2 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^2 \|_2^2 + w_3\| \mathfrak{a}_x^3 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^3 \|_\text{F}^2, 
\end{equation}  
where the explicit expressions of $\mathfrak{a}_x^1$, $\mathfrak{a}_x^2$ and $\mathfrak{a}_x^3$ are given in~\eqref{eq:ac_1_prolates},~\eqref{eq:ac_2_prolates} and~\eqref{eq:ac_3_prolates}, respectively.  
In the experiments, we set $w_1=w_2=w_3=1$. 
The LS problem was solved using Matlab's \texttt{lsqnonlin} solver for nonlinear LS problems. The expected $\hat \gamma$ in this experiment is one.

@@

The true volume used in the experiments in Figures~\ref{fig:cryo_detection} and~\ref{fig:BPTI_recon} was the Bovine Pancreatic Trypsin Inhibitor (BPTI) mutant with altered binding loop sequence, whos atomic model is available in the Protein Data Bank (PDB) %~\cite{pdb} 
as 1QLQ\footnote{\url{https://www.rcsb.org}}. We generated an EM map from this atomic model in UCSF Chimera~\cite{pettersen2004ucsf} at a resolution of 5~\AA, and cropped it to remove zeros at the boundary to obtain a volume of size $31^3$. For the experiment in Figure~\ref{fig:BPTI_recon}, the volume was downsampled to size $20^3$. For the experiment in Figure~\ref{fig:TRPV1_recon}, we used the TRPV1 in complex with DkTx and RTX, whose EM map is available in the  Electron Microscopy Data Bank (EMDB) %~\cite{emdb} 
as EMD-8117\footnote{\url{http://www.ebi.ac.uk/pdbe/emdb}}. The original map has size $192^3$, and was downsampled to size $20^3$. To generate the ground truth for our reconstructions, we expanded both volumes as in~\eqref{eq:volume_expansion_app} with cutoff $L_{\text{max}} = 5$ for TRPV1 and $L_{\text{max}} = 2$ for BPTI. For the TRPV1 reconstruction, the optimizer converged to a point giving relative $\ell_2$ error of $10^{-6}$ in the first three autocorrelations and an error of $10^{-1}$ in the expansion coefficients of the volume. For the BPTI reconstruction, the errors in the autocorrelations were $10^{-6}$, $10^{-7}$ and $10^{-8}$ for the third, second and first autocorrelations, respectively, while the error in the expansion coefficients of the volume was $5\times 10^{-2}$. This illustrates the ill-conditioning of the map between the volume and its first three autocorrelations that prevents us from obtaining \revised{high-resolution} results from noisy data.

The micrographs for the experiment presented in Figures~\ref{fig:cryo_detection} \revised{and~\ref{fig:cryo_noisy_micros_for_rec}} were generated as follows. We sample rotation matrices from SO(3) uniformly at random using the QR-based algorithm described in~\cite{stewart1980efficient}, and generate the projection of the volume corresponding to that rotation matrix using \verb|cryo_project| in ASPIRE\footnote{\url{http://spr.math.princeton.edu}}. %\cite{aspire}.
We keep track of the indices at which the upper left corner of a projection can be placed without violating the separation condition, so all projections are separated by at least \revised{$L-1$ pixels in each dimension, where  the projections are contained in a box of size $L\times L$.}  The location of the upper left corner of each new projection is picked uniformly at random from the set of available indices. We continue adding projections to the micrograph until no more projections can be added without violating the separation condition. \revised{In the experiments, we define SNR slightly differently than in the toy model, namely, as $\text{SNR} = \frac{\text{var}(\mathcal{I})}{\sigma^2}$ where $\text{var}(\mathcal{I})$ is the variance of our stack of micrographs and $\sigma^2$ is the variance of noise. The noise level in the noisy micrograph in Figure~\ref{fig:cryo_detection} was $\text{SNR} = 1/1024$, the noise level in Figure~\ref{fig:cryo_noisy_micros_for_rec} was $\text{SNR} = 1/16$ for the TRPV1 micrographs and $\text{SNR} = 1/64$ for the BPTI micrographs. We used 300 micrographs for each of the noisy reconstructions presented in Figure~\ref{fig:cryo_recon}.}

\revised{For the noisy reconstructions presented in Figure~\ref{fig:cryo_recon}, the LS problem~\eqref{eq:LS_cryo} was solved assuming spherical harmonic cutoffs of $L_{\text{max}} = 5$ for TRPV1 and $L_{\text{max}} = 2$ for BPTI, same as for the ground truth.}
For the experiment in Figure~\ref{fig:cryo_detection}, the LS problem~\eqref{eq:LS_cryo} was solved assuming the spherical harmonic cutoff for the volume is $L_{\text{max}}=0$, which is sufficient to recover a significant $\widehat\gamma$ in the presence of projections in the micrograph.

>>>>

Figure~\ref{fig:Einst_example} shows several estimated images for a growing number of micrographs. %, and a movie is available as supplementary material.
Figure~\ref{fig:error_per_micro} (see appendix) presents the normalized recovery error as a function of the amount of data available.  This is computed after fixing the reflection symmetries (see Section~\ref{sec:AC_analysis}). As evidenced by these figures, the ground truth image can be estimated increasingly well from increasingly many micrographs, without particle picking.

@@

\subsection{Cryo-EM} The application of autocorrelation analysis to cryo-EM follows the same mathematical principles.
The derivation of the first three autocorrelations of the micrographs and their relations to the volume itself are provided in Appendix~\ref{sec:ac_cryo}.
In particular, numerical evidence suggests that the third-order autocorrelation uniquely determines the 3-D volume. 
Figure~\ref{fig:cryo_recon} shows recovery of the 3-D volume from the clean autocorrelations {and from noisy micrographs.
Unfortunately, the mapping between the autocorrelations and volume seems to be ill-conditioned, preventing high-resolution recovery from noisy data.}
The details of the reconstruction algorithm are given in Appendix~\ref{sec:numeric_details}.
In the next section, we outline how we suggest to overcome the ill-conditioning in future work.

While we cannot provide a {high-resolution} 3-D reconstruction from noisy data with the current algorithm, our method can be easily applied to the problem of deciding whether a micrograph contains projections or merely pure noise---a problem considered in classical works in statistics~\cite{donoho2004higher} %and debates in the cryo-EM community~\cite{henderson2013avoiding}.
and cryo-EM~\cite{henderson2013avoiding}.
This task can be performed by considering solely the recovered $\gamma$ (the fraction of pixels occupied by projections in the micrograph), estimated as part of the recovery algorithm.
Figure~\ref{fig:cryo_detection} (see appendix) presents excerpts  of two noisy micrographs, only one of which contains projections.
In the presence of projections, the estimated $\gamma$ was $0.12$, corresponding to approximately $6784$ projections. 
On the other hand, the estimated $\gamma$ drops to $10^{-5}$ for the pure noise micrograph, corresponding to less than one projection.

\begin{figure}[h]
	\centering
	\begin{subfigure}[h]{0.45\textwidth}
		\centering
		\includegraphics[angle=270, width=.7\textwidth]{TRPV1_rec_incl_noisy.eps}
		\caption{TPRV1 with cutoff $L_{\text{max}}=5$}\label{fig:TRPV1_recon}
	\end{subfigure}
	\hfill
	\begin{subfigure}[h]{0.45\textwidth}
		\centering
		\includegraphics[angle=270, width=.7\textwidth]{BPTI_maxL2_L20.eps}
		\caption{BPTI with cutoff $L_{\text{max}}=2$}\label{fig:BPTI_recon}
	\end{subfigure} 
	\caption{\label{fig:cryo_recon} Reconstructions from the first, second and third order autocorrelations.
		The ground truth volumes were expanded according to~\eqref{eq:volume_expansion_app} with cutoff $L_{\text{max}}$.
		{The original molecules are shown in purple and the (smoothed) ground truths in blue  to illustrate the smoothing effect of our downsampling and truncation of the spherical harmonics expansion.
		The reconstructions  from clean autocorrelations are shown in yellow and recoveries from 
		autocorrelations estimated from noisy data in gray. For the noisy experiments, we used 300 micrographs with SNRs of 1/16 for TRPV1 and 1/64 for BPTI. We present excerpts from noisy micrographs side-by-side with the corresponding clean ones in Figure~\ref{fig:cryo_noisy_micros_for_rec} (see appendix)}.}
		 \vspace{-20pt} 	
\end{figure}
% -------------- Vertical subfigures: -------------------------------

@@

\begin{figure*}[t!]
	\centering
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{TRPV1_clean_micro.png}
		\caption{}
	\end{subfigure} \hfill %\hspace{2pt}
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{TRPV1_noisy_micro_SNR_16.png}
		\caption{}
	\end{subfigure} \hfill %\hspace{2pt}
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{clean_micro_BPTI.png}
		\caption{}
	\end{subfigure} \hfill
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{noisy_micro_BPTI_SNR_64.png}
		\caption{}
	\end{subfigure}
	\caption{\label{fig:cryo_noisy_micros_for_rec}  {Excerpts from the noisy micrographs used for the reconstructions in Figure~\ref{fig:cryo_recon} and the corresponding clean excerpts. (a) Excerpt of the clean TRPV1 micrograph; (b) Excerpt of the noisy TRPV1 micrograph with SNR $= 1/16$; (c) Excerpt from the clean BPTI micrograph; (d) Excerpt from the noisy BPTI micrograph with SNR $= 1/64$.}} 
\end{figure*}

\begin{figure*}[t!]
	\centering
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\includegraphics[scale=0.35]{TRPV1_trunc_FSC.eps}
		\caption{}
	\end{subfigure} \hfill %\hspace{2pt}
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\includegraphics[scale=0.35]{BPTI_FSC.eps}
		\caption{}
	\end{subfigure} 
	\caption{\label{fig:cryo_FSC}  {Fourier Shell Correlations (FSCs) for the noisy reconstructions presented in Figure~\ref{fig:cryo_recon} using the $0.5$ criterion, compared to the (smoothed) ground truth. (a) FSC for the TRPV1 reconstruction, giving resolution of $24 \AA$; (b) FSC for the BPTI reconstruction, giving resolution of $13 \AA$.}} 
\end{figure*}

@@

For the {clean} experiments in Figure~\ref{fig:cryo_recon}, we computed the analytical first three autocorrelations of the volume as explained in Appendix~\ref{sec:ac_cryo}.
To estimate the coefficients of the volume itself, we solve the LS problem
\begin{equation} \label{eq:LS_cryo}
\min_{\hat{x},\hat{\gamma}} w_1 \vert \mathfrak{a}_x^1 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^1 \vert^2 + w_2\| \mathfrak{a}_x^2 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^2 \|_2^2 + w_3\| \mathfrak{a}_x^3 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^3 \|_\text{F}^2, 
\end{equation}  
where the explicit expressions of $\mathfrak{a}_x^1$, $\mathfrak{a}_x^2$ and $\mathfrak{a}_x^3$ are given in~\eqref{eq:ac_1_prolates},~\eqref{eq:ac_2_prolates} and~\eqref{eq:ac_3_prolates}, respectively.  
In the experiments, we set $w_1=w_2=w_3=1$. 
The LS problem was solved using Matlab's \texttt{lsqnonlin} solver for nonlinear LS problems. The expected $\hat \gamma$ in this experiment is one.

@@

The true volume used in the experiments in Figures~\ref{fig:cryo_detection} and~\ref{fig:BPTI_recon} was the Bovine Pancreatic Trypsin Inhibitor (BPTI) mutant with altered binding loop sequence, whos atomic model is available in the Protein Data Bank (PDB) %~\cite{pdb} 
as 1QLQ\footnote{\url{https://www.rcsb.org}}. We generated an EM map from this atomic model in UCSF Chimera~\cite{pettersen2004ucsf} at a resolution of 5~\AA, and cropped it to remove zeros at the boundary to obtain a volume of size $31^3$. For the experiment in Figure~\ref{fig:BPTI_recon}, the volume was downsampled to size $20^3$. For the experiment in Figure~\ref{fig:TRPV1_recon}, we used the TRPV1 in complex with DkTx and RTX, whose EM map is available in the  Electron Microscopy Data Bank (EMDB) %~\cite{emdb} 
as EMD-8117\footnote{\url{http://www.ebi.ac.uk/pdbe/emdb}}. The original map has size $192^3$, and was downsampled to size $20^3$. To generate the ground truth for our reconstructions, we expanded both volumes as in~\eqref{eq:volume_expansion_app} with cutoff $L_{\text{max}} = 5$ for TRPV1 and $L_{\text{max}} = 2$ for BPTI. For the TRPV1 reconstruction, the optimizer converged to a point giving relative $\ell_2$ error of $10^{-6}$ in the first three autocorrelations and an error of $10^{-1}$ in the expansion coefficients of the volume. For the BPTI reconstruction, the errors in the autocorrelations were $10^{-6}$, $10^{-7}$ and $10^{-8}$ for the third, second and first autocorrelations, respectively, while the error in the expansion coefficients of the volume was $5\times 10^{-2}$. This illustrates the ill-conditioning of the map between the volume and its first three autocorrelations that prevents us from obtaining {high-resolution} results from noisy data.

The micrographs for the experiments presented in Figures~\ref{fig:cryo_detection}  {and~\ref{fig:cryo_noisy_micros_for_rec}} were generated as follows. 
We sample rotation matrices from SO(3) uniformly at random using the QR-based algorithm described in~\cite{stewart1980efficient}, and generate the projection of the volume corresponding to that rotation matrix using \verb|cryo_project| in ASPIRE\footnote{\url{http://spr.math.princeton.edu}}. %\cite{aspire}.
The projections for the experiments in Figure~\ref{fig:cryo_noisy_micros_for_rec} were obtained from the smoothed volumes, not the original ones. 
We keep track of the indices at which the upper left corner of a projection can be placed without violating the separation condition, so all projections are separated by at least {$L-1$ pixels in each dimension, where  the projections are contained in a box of size $L\times L$.}  The location of the upper left corner of each new projection is picked uniformly at random from the set of available indices. We continue adding projections to the micrograph until no more projections can be added without violating the separation condition. {In the experiments, we define SNR slightly differently than in the toy model, namely, as $\text{SNR} = \frac{\text{var}(\mathcal{I})}{\sigma^2}$ where $\text{var}(\mathcal{I})$ is the variance of our stack of micrographs and $\sigma^2$ is the variance of noise. The noise level in the noisy micrograph in Figure~\ref{fig:cryo_detection} was $\text{SNR} = 1/1024$, the noise level in Figure~\ref{fig:cryo_noisy_micros_for_rec} was $\text{SNR} = 1/16$ for the TRPV1 micrographs and $\text{SNR} = 1/64$ for the BPTI micrographs. We used 300 micrographs for each of the noisy reconstructions presented in Figure~\ref{fig:cryo_recon}.}

{For the noisy reconstructions presented in Figure~\ref{fig:cryo_recon}, the LS problem~\eqref{eq:LS_cryo} was solved assuming spherical harmonic cutoffs of $L_{\text{max}} = 5$ for TRPV1 and $L_{\text{max}} = 2$ for BPTI, same as for the ground truth.}
For the experiment in Figure~\ref{fig:cryo_detection}, the LS problem~\eqref{eq:LS_cryo} was solved assuming the spherical harmonic cutoff for the volume is $L_{\text{max}}=0$, which is sufficient to recover a significant $\widehat\gamma$ in the presence of projections in the micrograph.
